Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-21 manuscript.rmd
1. Changed title 2. TRied to add para first line indentation 2. multi-line equation made to work with word 3. other minor formatting

---
title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

# Introduction
\
    Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

@@

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

@@

\begin{align*}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\
+ RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{align*}

@@

\begin{equation}
K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs. $\\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *p*-values are given in inset.

@@

Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama). Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively. *p*-values are given only when they are $\\ge$ 0.05. Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.

@@

```{r eddy, results = 'asis', echo = FALSE}
knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes. w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed. The bar indicates 30-min averaging and the apex indicates fluctuation ($\\overline{x` = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
```

>>>>

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

# Introduction

\     
Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

@@

\
|   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

@@

<!-- # For html -->
<!-- \begin{align*} -->
<!-- \overline{O} = \frac{1}{6}\sum RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\ -->
<!-- + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m}) -->
<!-- (\#eq:obj-func) -->
<!-- \end{align*} -->

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) \\
& + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m})\right)
(\#eq:obj-func)
\end{split}
\end{equation}

@@

\begin{equation}
K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs. $\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *p*-values are given in inset.

@@

Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama). Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively. *p*-values are given only when they are $\ge$ 0.05. Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\Psi$ in MPa.

@@

```{r eddy, results = 'asis', echo = FALSE}
knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes. w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed. The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
```
